Neural and Symbolic Spatiotemporal Optimization of Cosmological Particle-Mesh Simulation

### 2023-05-29 docs/papers/sto/sto.tex
FIXUP draft: loss engineering

A mean squared error (MSE) loss on the fields is dominated by the small
scale modes \citep{HeEtAl2019}.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
Our loss on a field $f$ sums up its MSE relative to the target field
power in logarithmic bins:
%
\begin{equation}
\cL_f = \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk) \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2},
\end{equation}
%
\textbf{
where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
factor that suppresses the high $k$ contribution in Eulerian fields such
as the density field (with $w=1$ for the Lagrangian displacement field).
This is because a PM solver gradually loses small scale information step
by step, and without $w$ setting the priority, the loss function could
encourage the network to fit the small scales at the cost of the
intermediate scales.
As a result, we can set the loss mesh size the same as that of the PM
force solver.
We can try for $w$:
%
\begin{equation}
\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{and its square},
\prod_{j=1}^3 \cos (k_j l / 2) \textrm{and its square},
\end{equation}
%
which vanish when any $k_j$ reaches the Nyquist.
}
\YL{Maybe add a figure of $w$ along axis, face diagonal and body
digonal, if this works.}

To understand the form of the loss function, let's introduce for each
mode the square root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
\end{equation}
%
and the cross correlation coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
\end{equation}
%
Now, under the mild assumptions that $T$ and $r$ are smooth and
isotropic, we can show
%
\begin{equation}
\cL_f \propto \int \!\d\ln\!k\, w(\vk)
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
\end{equation}
%
that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
when $\hat f$ is identical to $f$.
Above we choose the binned version instead for the robustness against
the noisy modewise $|f(\vk)|^2$ in the denominator.
The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
loss, while our modification is important to balance among all scales so
that the neural network does not focus too much on either high or low
$k$'s.

>>>>

A mean squared error (MSE) loss on the fields is dominated by the small
scale modes \citep{HeEtAl2019}.
In this work we choose uniform weights in the scale space on the
relative MSE.
We bin the modes logarithmically by \nicefrac13 octave in scale, with
the $i$-th $k$-bin $K_i \triangleq [k_i, k_{i+1})$, where $k_i =
2^{1+i/3} \pi / L$.
Our loss on a field $f$ sums up its MSE relative to the target field
power in logarithmic bins:
%
\begin{equation}
\cL_f = \frac1{N_K} \sum_i \ln \biggl[
\frac{\sum_{k \in K_i} \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2} \biggr],
\end{equation}
%
\begin{equation}
\cL_f = \frac1{N_K} \sum_i
\frac{\sum_{k \in K_i} w(\vk) \bigl| \hat f(\vk) - f(\vk) \bigr|^2}
     {\sum_{k' \in K_i} |f(\vk')|^2},
\end{equation}
%
where $N_K$ is the number of $k$-bins and $w(\vk)$ is a nonnegative
factor that suppresses the high $k$ contribution in Eulerian fields such
as the density field (with $w=1$ for the Lagrangian displacement field).
This is because a PM solver gradually loses small scale information step
by step, and without $w$ setting the priority, the loss function could
encourage the network to fit the small scales at the cost of the
intermediate scales.
As a result, we can set the loss mesh size the same as that of the PM
force solver.
We can try for $w$:
%
\begin{equation}
\prod_{j=1}^3 \frac{k_j l / 2}{\tan(k_j l /2)} \textrm{to some power},
\prod_{j=1}^3 \cos (k_j l / 2) \textrm{to some power},
\end{equation}
%
which vanish when any $k_j$ reaches the Nyquist.
\YL{Maybe add a figure of $w$ along axis, face diagonal and body
digonal, if this works.}

To understand the form of the loss function, let's introduce for each
mode the square root ratio of power
%
\begin{equation}
T(\vk) \triangleq
\sqrt{\frac{|\hat f(\vk)|^2}{|f(\vk')|^2}},
\end{equation}
%
and the cross correlation coefficient
%
\begin{equation}
r(\vk) \triangleq
\frac{\Re \bigl[ \hat f(\vk) f^*(\vk) \bigr]}
     {\sqrt{|\hat f(\vk)|^2 |f(\vk')|^2}}.
\end{equation}
%
Now, under the mild assumptions that $T$ and $r$ are smooth and
isotropic, we can show
%
\begin{equation}
\cL_f \propto \int \!\d\ln\!k\, \ln
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq -\infty;
\end{equation}
%
\begin{equation}
\cL_f \propto \int \!\d\ln\!k\, w(\vk)
\Bigl\{ \bigl[ 1 - T(\vk) \bigr]^2
  + 2 T(\vk) \bigl[ 1 - r(\vk) \bigr] \Bigr\} \geq 0,
\end{equation}
%
that only vanishes when both $T$ and $r$ are unity for all modes, i.e.,
when $\hat f$ is identical to $f$.
Above we choose the binned version instead for the robustness against
the noisy modewise $|f(\vk)|^2$ in the denominator.
The derivation here is similar to that in \citet{HeEtAl2019} on the MSE
loss, while our modification is important to balance among all scales so
that the neural network does not focus too much on either high or low
$k$'s.
